Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/relatedwork.tex
Quick pass.

As Jorring and Scherlis argue frequency reduction and precomputation
are common techniques for both designing algorithms and performing
compiler optimizations~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  For example, dynamic programming and loop hoisting,
as well as the work presented here are a frequency reduction
techniques.
%
The idea behind precomputation is to identify computations that can be
performed earlier, for example at compile time if their inputs are
available statically and perform them at that earlier time.  For
example, dynamic algorithms, partial evaluation, and incremental
computation are all precomputation techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of its static inputs, which are the subset of the inputs that
become available at compilation time, partial evaluation generates a
specialized program by performing the computations that depend on the
statics inputs~\cite{jones96}.  The roots of partial evaluation go
back to Kleene, who in 1952 proved that a multivariate program
(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
however.  Lombardi appears to be the first person to use the term in
his work on incremental computation~\cite{Lombardi67}.  Since then,
there has a lot of work on partial evaluation, only broad-strokes of
which we can mention in this here.  We refer the interested reader to
the excellent book by Jones, Gommard, and Sestoft for a comprehesive
discussion of work until early 90's~\cite{JGS93}.

At a high level, earlier approaches to partial evaluation can be
viewed as operating in two stages: binding time analysis and program
specialization.  For a multivariate program with clearly marked static
and dynamic arguments, binding-time analysis identifies all the parts
of the program that can be computed by the knowledge of static
arguments. Using this information and the static arguments themselves,
program specialization specializes the program to a partially
evaluated program that can be used with many different dynamic
arguments.  This approach has been applied to construct partial
evaluators for a number of languages such as
Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
part first-order subsets.

@@

The type-system component of our work is closely related to this later
line of work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose.  The key difference between our work and this prior work is
the staging or the splitting algorithm, which given a two-staged
program splits it into two stages that can be executed sequentially,
passing information from the earlier to the latter stage by
construction. 
%
Unlike in partial evaluation, our splitting algorithm requires no
knowledge of the arguments themselves. The only information that is
required is their staging. Using just the staging information, the
splitting algorithm can generate programs that can perform non-trivial
transformations on the code automatically.  As our examples illustrate
such staging transformations can improve the efficiency of the source
program, sometimes in asymptotically significant ways.  
We are not aware of prior work in the context of higher-order
functional languages that presents such a splitting algorithm.

An important class of pre-computation techniques include incremental
computation, where a program can respond to small changes to its input
efficiently by recomputing only the parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
As other pre-computation techniques, incremental computation can be
thought as a staged model, where changes (inputs) may come over a
period of time in stages. 
%
It does not, however, fix any of the inputs or anticipate the future
inputs, allowing in the most general case for all data to change. As a
result, its benefits are dependent on the data instance (nature of the
inputs). In addition, incremental computation techniques typically
utilize significant run-time (dynamic) information, because no
assumptions about input instances are made.  For example, we can apply
incremental computation to our quickselect example
(ref{sec:overview}), but we would not be able to achieve the bounds
that we did because incremental computation techniques would unfold
the quickselect function based on the demanded ranks, which can
require linear time (as opposed to logarithmic) at each stage.  In
addition, such an implementation would maintain rather sophisticated
data structures dynamically at run time, the benefits of which depend
highly on the forms of changes applied.

\paragraph{Frequency-reduction techniques.}
Perhaps the closest work to ours is algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  As our
work, Knoblock and Ruf are concerned about the problem of splitting an
explicitly staged program into two stages.  However, they only
consider, simple first-order language and straight-line
programs. Automatic pass seperation has also been a topic of interest
in programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Seperation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass seperation to the compiler. However, all
prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
limited to simple, imperative languages.

>>>>

Frequency reduction and precomputation are common techniques for both
designing algorithms and performing compiler
optimizations~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  Dynamic programming, loop hoisting, and pass
separation (presented here) are all examples of frequency reduction
techniques.
%
The idea behind precomputation is to identify computations that can be
performed earlier (e.g., at compile time) if their inputs are
available statically and perform them at that earlier time. Dynamic
algorithms, partial evaluation, and incremental computation are all
examples of precomputation techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of static inputs (inputs that are available at compilation
time), partial evaluation generates a specialized program by
performing the computations that depend on the statics
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene, who in 1952 proved that a multivariate program (function)
$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
Kleene did not intend to improve efficiency, however.  Lombardi
appears to be the first person to use the term in his work on
incremental computation~\cite{Lombardi67}.  Since then, there has a
lot of work on partial evaluation, only broad-strokes of which we can
mention in this here.  We refer the interested reader to the excellent
book by Jones, Gommard, and Sestoft for a comprehesive discussion of
work until early 90's~\cite{JGS93}.

Earlier approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the static arguments themselves, program
specialization specializes the program to a partially evaluated
program that can be used with many different dynamic arguments.  This
approach has been applied to construct partial evaluators for a number
of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism},
considering for the most part first-order subsets.

@@

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose.  The key difference between our work and this prior work is
our focus on the problem of pass separation: we developed an algorithm
for splitting terms in the proposed staged language.
%the staging or the splitting algorithm, which given a two-staged
%program splits it into two stages that can be executed sequentially,
%passing information from the earlier to the latter stage by
%construction.
%
Unlike partial evaluation, splitting requires no knowledge of program
argument values, only their staging.
%Using just the staging
%information, the splitting algorithm can generate programs that can
%perform non-trivial transformations on the code automatically.  As our
%examples illustrate such staging transformations can improve the
%efficiency of the source program, sometimes in asymptotically
%significant ways.
We are not aware of prior work in the context of
higher-order functional languages that presents such a splitting
algorithm.

An important class of pre-computation techniques include incremental
computation, where a program can respond to small changes to its input
efficiently by recomputing only the parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Like other pre-computation techniques, incremental computation can be
thought of as a staged model, where changes (in inputs) come over a
period of time in stages.
%
However, incremental computation does not fix any of its inputs or anticipate the future
inputs, allowing in the most general case for all data to change. As a
result, its benefits are dependent on the data instance (nature of the
inputs). In addition, incremental computation techniques typically
utilize significant run-time (dynamic) information, because no
assumptions about input instances are made.  For example, we can apply
incremental computation to our quickselect example
(ref{sec:overview}), but we would not be able to achieve the bounds
that we did because incremental computation techniques would unfold
the quickselect function based on the demanded ranks, which can
require linear time (as opposed to logarithmic) at each stage.  In
addition, such an implementation would maintain rather sophisticated
data structures dynamically at run time, the benefits of which depend
highly on the forms of changes applied.

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
our work, Knoblock and Ruf are concerned about the problem of
statically splitting an explicitly staged program into two stages.
However, they only consider simple first-order language and
straight-line programs and also treat all computations guarded by
second stage conditionals as second stage computations.

Automatic pass seperation has also been a topic of interest in
programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Seperation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass seperation to the compiler. However, all
prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
limited to simple, imperative languages.
